---
title: "SLBP analysis"
author: "Sudeep Sahadevan, Thomas Schwarzl and Thileepan Sekaran"
date: "3/12/2021"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 5
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This markdown file contains the complete analysis of SLBP example dataset and follows the same analysis steps and file naming patterns as it is described in the chapter. 

## Installation of DEWSeq and setup

```{r install, eval=FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("DEWSeq")
```

## Load  libraries

```{r libs}
suppressPackageStartupMessages(require(DEWSeq)) 
suppressPackageStartupMessages(require(IHW))
suppressPackageStartupMessages(require(tidyverse))
suppressPackageStartupMessages(require(data.table))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(ggrepel))

```

`suppressPackageStartupMessages` ignores all start up messages

## Data import

```{r wd, eval=FALSE}
setwd('/path/to/SLBP_analysis')
```
```{r countMat}
count_matrix <- fread('counts/SLBP_K562_w50s20_counts.txt.gz', stringsAsFactors=FALSE, sep="\t", header=TRUE)
count_matrix <- column_to_rownames(count_matrix,'unique_id')
head(count_matrix)
```
```{r sampleMat}
col_data <- data.frame(type=c('IP','IP','SMI'), row.names=colnames(count_matrix))
head(col_data)
```
```{r annotationFile}
annotation_file <- 'annotation/SLBP_K562_w50s20_annotation.txt.gz'
```

### Create DESeqDataset
```{r ddw1}
ddw <- DESeqDataSetFromSlidingWindows(countData=count_matrix, colData=col_data, annotObj=annotation_file, design=~type)
```
```{r ddw2}
head(ddw)
```

## Estimate size factors 

```{r sf1}
ddw <- estimateSizeFactors(ddw)
sizeFactors(ddw)
```

### Estimate size factors for only protein coding genes

```{r sf2}
ddw_mRNAs <- ddw[ rowData(ddw)[,"gene_type"] == "protein_coding", ]
ddw_mRNAs <- estimateSizeFactors(ddw_mRNAs)
sizeFactors(ddw) <- sizeFactors(ddw_mRNAs)
```

```{r sf2_1}
sizeFactors(ddw_mRNAs)
```
### Estimate size factors without significant windows

In this context, the aim is to use only the windows that which do not show a large difference between IP and SMI samples for normalization.

This step uses `local` fit for dispersion estimation and Likelihood-ratio test (LRT)

```{r sf3_1}
ddw_tmp <- ddw
ddw_tmp <- estimateDispersions(ddw_tmp, fitType = "local", quiet = TRUE)
ddw_tmp <- nbinomLRT(ddw_tmp,full = ~type,reduced = ~1)
tmp_significant_windows <- 
                results(ddw_tmp,
                    contrast = c("type", "IP", "SMI"),
                    tidy = TRUE,
                    filterFun = ihw) %>% 
                dplyr::filter(padj < 0.05) %>% 
                .[["row"]]
rm(ddw_tmp)
```

Now estimate the size factors without the significant windows.

```{r sf3_2}
ddw_mRNAs <- ddw_mRNAs[ !rownames(ddw_mRNAs) %in% tmp_significant_windows, ]
ddw_mRNAs <- estimateSizeFactors(ddw_mRNAs)
sizeFactors(ddw) <- sizeFactors(ddw_mRNAs)
```

```{r sf3_3}
sizeFactors(ddw_mRNAs)
rm(ddw_mRNAs)
```
## Prefiltering

```{r pref}
keep <- rowSums(counts(ddw)) >= 1
ddw <- ddw[keep,]
ddw
```

## Estimate dispersion

By default, DESeq2 uses `parametric` fit for dispersion estimation

```{r parametric, eval=FALSE}
ddw <- estimateDispersions(ddw,quiet=TRUE)
```

Users can also opt for `local` dispersion estimation

```{r local, eval=FALSE}
ddw <- estimateDispersions(ddw,fitType='local',quiet=TRUE)
```

### Decide fit

Instead of picking a fit type at random, it is possible to decide the best fit based on the residuals after calculating the dispersion estimates with each type. This step is based on a [bioconductor post](https://support.bioconductor.org/p/81094/)

```{r decideFit}
ddw_param <- estimateDispersions(ddw,quiet=TRUE)
ddw_local <- estimateDispersions(ddw,fitType='local',quiet=TRUE)
```

```{r paramPlot,fig.width=8}
plotDispEsts(ddw_param)
```

```{r localPlot,fig.width=8}
plotDispEsts(ddw_local)
```





```{r resids}
parametric_resid <- na.omit(with(mcols(ddw_param),abs(log(dispGeneEst)-log(dispFit))))
local_resid <- na.omit(with(mcols(ddw_local),abs(log(dispGeneEst)-log(dispFit))))
```

```{r summary1}
summary(parametric_resid)
```
```{r summary2}
summary(local_resid)
```
```{r residHistogram, fig.width=8}
resid_df <- data.frame(residuals=c(parametric_resid,local_resid),fitType=c(rep("parametric",length(parametric_resid)),rep("local",length(local_resid))))
ggplot(resid_df, aes(x = residuals, fill = fitType)) + scale_fill_manual(values = c("darkred", "darkblue")) + geom_histogram(alpha = 0.5, position='identity', bins = 100) + theme_bw()
```

The residual summaries and histogram clearly indicates that the `local` fit produces a better fit, hence smaller residuals, and therefore we proceed with using local fit for the rest of the analysis.

```{r chooseFit}
ddw <- ddw_local
```

this can also be done just based on the median values as:

```{r chooseFit_auto, eval=FALSE}
if (median(local_resid) <= median(parametric_resid)){
    cat("chosen fitType: local")
    ddw <- ddw_local
  }else{
    cat("chosen fitType: parametric")
    ddw <- ddw_param
  }
```

```{r removeOb}
rm(ddw_local,ddw_param,resid_df,parametric_resid,local_resid)
```
## Differential expressed windows analysis

Using LRT

```{r LRT}
ddw <- nbinomLRT(ddw, full = ~type, reduced = ~1)
```

Using Wald test

```{r Wald, eval=FALSE}
ddw <-  nbinomWaldTest(ddw)
```

### Extract significant windows 

```{r sigWindows}
resultWindows <- resultsDEWSeq(ddw, contrast = c("type", "IP", "SMI"),tidy = TRUE) %>% as_tibble
resultWindows
```

## Multiple hypothesis correction

### FDR correction using BH

```{r bh}
resultWindows[,'p_adj'] <- p.adjust(resultWindows$pvalue, method="BH")
```

### Using IHW

```{r ihw}
resultWindows[,"p_adj_IHW"] <- adj_pvalues(ihw(pvalue ~ baseMean, data = resultWindows, alpha = 0.05, nfolds = 10))
```

## Combining windows to regions

```{r regions,message=FALSE}
resultRegions <- extractRegions(windowRes=resultWindows, padjCol="p_adj_IHW", padjThresh = 0.05, log2FoldChangeThresh=1) %>% as_tibble
resultRegions
```

## Exporting results

```{r toBEd}
toBED(windowRes=resultWindows, regionRes=resultRegions, padjThresh=0.05, padjCol="p_adj_IHW", fileName="SLBP_regions_w50s20.bed")
```

## Visualization

An R function to visualize enriched regions as volcano plots is available in [this github repo](https://github.com/EMBL-Hentze-group/DEWSeq_analysis_helpers). Please refer to README files in this repo for additional details.

```{r volcano, fig.width=8}
source('https://raw.githubusercontent.com/EMBL-Hentze-group/DEWSeq_analysis_helpers/master/Volcano_plot/volcanoplot.R')
# subset of all results
testdf <- resultWindows[ resultWindows$p_adj_IHW<=0.95, ]
volcanoplot(testdf,padj_col = 'p_adj_IHW',gene_names = c('HIST1H4K','HIST1H2AK','HIST1H1B','HIST1H3J','HIST1H2AM','H2AFX'))
```

